---
title: "Naomi_Torrisi_Comp_Genomics_CW"
author: "Naomi Torrisi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install("rtracklayer")

library(rtracklayer)
library(tidyverse)

```


SAQ 1: Briefly describe the characteristics of the cancer genome (10 marks)
```{r}

```


SAQ 2: 
1.	What is cancer immunotherapy? (2 marks)
```{r}

```

2.	Give two examples of cancer immunotherapy, and also brief describe the underlying mechanism of each immunotherapy? (4 marks)
```{r}

```

3.	Briefly describe strategies that tumour cells take to evade immune surveillance. (4 marks) 
```{r}

```

# LAQ 1:
LAQ 1:
High grade serous ovarian cancer cell lines Kuramochi and Snu119 were single cell sorted, such that one individual cell was dispensed into one well of a 96 well plate to form a new clonal population (a ‘clone’). Each new clone was expanded into a new clonal population over 22 cell division cycles. Single cell DNA sequencing (low pass whole genome sequencing) was performed on small subsets of (i) the parental population that the single cell was derived from, and (ii) the new clonal populations (C8 & E10 (Kuramochi) and A12 & F12 (Snu119)). (Total 25 marks)

Get the data required for this question from ‘/data/teaching/bci_teaching/Computational_genomics/’. You can find further instructions on how to transfer the data to your hpc folder in the slide no 8 in ‘Practical 2_Study Tumour Evolution using single cell sequencing_2026.pptx’ file.

1.	5 marks are available for successfully running the data through Aneufinder and MEDICC2 and selecting the relevant data to contribute to the following questions. 

```{r}
## code to process data using Aneufinder:
# AneuFinder was run in RStudio OnDemand / Apocrita
# Aneufinder and Medicc2 folders copied over to home directory from shared bci teaching Computational_genomics folder:

## on terminal:
## access bci teaching folder
#cd /data/teaching/bci_teaching/Computational_genomics
## check for aneufinder and medicc2 folders
#ls

##copy data into home directory
#cp -rf AneuFinder /data/home/ha25679
#cp -rf Medicc2 /data/home/ha25679

## check files are in home directory
#cd /data/home/ha25679
#ls

## you should now see:
#AneuFinder
#Medicc2

## In console:

#if (!require("BiocManager", quietly = TRUE))
#install.packages("BiocManager")

#install.packages ("gplots")
#install.packages("Rcpp")
#install.packages("https://cran.r-project.org/src/contrib/Archive/ReorderCluster/ReorderCluster_2.0.tar.gz", repos=NULL, type="source")
#BiocManager::install("BSgenome.Hsapiens.UCSC.hg38")
#BiocManager::install("Aneufinder")

#setwd(“/data/home/ha25679/AneuFinder)

# Key parameters: hg38 assembly, GC correction, 1 Mb bins, edivisive segmentation.

## load libraries
#library(Aneufinder)
#library(BSgenome.Hsapiens.UCSC.hg38)

#Aneufinder(
#  inputfolder = "Kuramochi_parent/",
#  outputfolder = "Kuramochi_parent/Aneufinder/",
#  assembly = "hg38",
#  numCPU = 2,
#  binsizes = 1e6,
#  blacklist = "hg38-blacklist.v2.bed",
#  correction.method = "GC",
#  GC.BSgenome = BSgenome.Hsapiens.UCSC.hg38,
#  refine.breakpoints = FALSE,
#  method = "edivisive",
#  chromosomes = c("1","2","3","4","5","6","7","8","9","10",
#                  "11","12","13","14","15","16","17","18","19",
#                  "20","21","22","X")
#)

## view the plots
#files=list.files('Kuramochi_E10/Aneufinder/MODELS/method-edivisive/', full.names=TRUE)
#models = loadFromFiles(files)
#plot(files[2], type='profile')


## Single-cell copy-number profiles were generated using AneuFinder, and tumour evolutionary relationships were reconstructed using MEDICC2.
## From AneuFinder, genome-wide single-cell copy-number profiles and karyograms were selected for downstream analysis of copy-number alterations, breakpoints and CNA sizes.

# Define directory
pdf_dir <- "/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Kuramochi_parent/LAQ1_Kuramochi_parent_anneufinder_plots"

# Define PDF paths
pdf_paths <- c(
  file.path(pdf_dir, "genomeHeatmap_binsize_1e+06_stepsize_1e+06.pdf"),
  file.path(pdf_dir, "aneuploidyHeatmap_binsize_1e+06_stepsize_1e+06.pdf"),
  file.path(pdf_dir, "karyograms_binsize_1e+06_stepsize_1e+06.pdf"),
  file.path(pdf_dir, "profiles_binsize_1e+06_stepsize_1e+06.pdf")
)

# Check files exist
file.exists(pdf_paths)

# Load pdftools
library(pdftools)

# Convert each PDF to PNG (page 1 only)
png_files <- character(length(pdf_paths))

for (i in seq_along(pdf_paths)) {
  
  png_files[i] <- file.path(tempdir(), paste0("plot_", i, ".png"))
  
  pdftools::pdf_convert(
    pdf_paths[i],
    format = "png",
    pages = 1,
    filenames = png_files[i]
  )
}

# Include in knitted output
knitr::include_graphics(png_files)



############################ REWORD CHAT GPT MARK #########################################################################################################################################################################################################################################################In addition to single-cell CNV BED files, AneuFinder generates a genome-wide breakpoint hotspot density track (WIG format), summarising recurrent breakpoint locations across cells, which was used to assess whether CNAs are evenly distributed across chromosomes.

```
```{r warning=FALSE}
## MEDICC2 outputs: phylogenetic tree and branch-length metrics
### code to process data using Aneufinder:
# MEDICC2 was run in RStudio OnDemand / Apocrita
## In terminal:

## go to Medicc2 folder
# cd /data/home/ha25679/Medicc2

## load conda (conda already available, unable to locate modulefile for 'anaconda 3')
#conda create -n medicc python=3.9
#conda activate medicc
#conda install -c bioconda -c conda-forge medicc2
#mkdir medicc_output
#ls
#medicc2 mcs_Kuramochi_E10_1000kb.tsv medicc_output/

## List all files in this cell (Kuramochi_parent)
list.files(
  path = "/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Kuramochi_parent/LAQ1_Kuramochi_parent_Medicc2_plots",
  pattern = "mcs_Kuramochi_E10_1000kb"
)

# set file path
setwd("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Kuramochi_parent/LAQ1_Kuramochi_parent_Medicc2_plots")

## check MEDICC2 files are present
list.files()

## open MEDICC2 phylogenetic tree
browseURL("mcs_Kuramochi_parent_1000kb_final_tree.png")

## open copy-number profiles used by MEDICC2
browseURL("mcs_Kuramochi_parent_1000kb_cn_profiles.pdf")

## read branch-length table (used later for Q5)
branch_lengths <- read.delim("mcs_Kuramochi_parent_1000kb_branch_lengths.tsv", header = TRUE)

## check branch lengths
head(branch_lengths)

```
### Load in breakpoint files for part 3
```{r warning=FALSE}
## read in bedframes for all clones
library(tidyverse)
all_breakpoint_files <- list(
  breakpoint_snu119_parent = as.data.frame(
    import("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Bedfiles/Snu119_parent_bedfiles/binsize_1e+06_stepsize_1e+06_breakpoints.bed.gz")
  ),
  breakpoint_snu119_A12 = as.data.frame(
    import("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Bedfiles/Snu119_A12_bedfiles/binsize_1e+06_stepsize_1e+06_breakpoints.bed.gz")
  ),
  breakpoint_snu119_F12 = as.data.frame(
    import("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Bedfiles/Snu119_F12_bedfiles/binsize_1e+06_stepsize_1e+06_breakpoints.bed.gz")
  ),
  breakpoint_kuramochi_parent = as.data.frame(
    import("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Bedfiles/Kuramochi_parent_bedfiles/binsize_1e+06_stepsize_1e+06_breakpoints.bed.gz")
  ),
  breakpoint_kuramochi_E10 = as.data.frame(
    import("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Bedfiles/E10_bedfiles/binsize_1e+06_stepsize_1e+06_breakpoints.bed.gz")
  ),
  breakpoint_kuramochi_C8 = as.data.frame(
    import("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Bedfiles/C8_bedfiles/binsize_1e+06_stepsize_1e+06_breakpoints.bed.gz")
  )
)

## Add bed file name to end of each dataframe in the list
all_breakpoint_files <- setNames(
  lapply(names(all_breakpoint_files), function(x) {
    
    df <- all_breakpoint_files[[x]]
    
    names(df)[1] <- "chromosome"
    df$bedfile <- x
    
    df
    
  }),
  names(all_breakpoint_files)
)

## View bedfiles as dataframes
all_breakpoint_files

```

### Load in cnv bed files for part 4
```{r warning=FALSE}

## load libraries
library(rtracklayer)
library(dplyr)

## import cnv bedfiles
all_cnv_files <- list(
  cnv_snu119_parent = as.data.frame(
    import("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Bedfiles/Snu119_parent_bedfiles/binsize_1e+06_stepsize_1e+06_CNV.bed.gz")
  ),
  cnv_snu119_A12 = as.data.frame(
    import("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Bedfiles/Snu119_A12_bedfiles/binsize_1e+06_stepsize_1e+06_CNV.bed.gz")
  ),
  cnv_snu119_F12 = as.data.frame(
    import("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Bedfiles/Snu119_F12_bedfiles/binsize_1e+06_stepsize_1e+06_CNV.bed.gz")
  ),
  cnv_kuramochi_parent = as.data.frame(
    import("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Bedfiles/Kuramochi_parent_bedfiles/binsize_1e+06_stepsize_1e+06_CNV.bed.gz")
  ),
  cnv_kuramochi_E10 = as.data.frame(
    import("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Bedfiles/E10_bedfiles/binsize_1e+06_stepsize_1e+06_CNV.bed.gz")
  ),
  cnv_kuramochi_C8 = as.data.frame(
    import("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Bedfiles/C8_bedfiles/binsize_1e+06_stepsize_1e+06_CNV.bed.gz")
  )
)

## Add bed file name to end of each dataframe in the list
all_cnv_files <- setNames(
  lapply(names(all_cnv_files), function(x) {
    
    df <- all_cnv_files[[x]]
    
    names(df)[3] <- "chromosome"
    df$bedfile <- x
    
    df
    
  }),
  names(all_cnv_files)
)

## View bedfiles as dataframes
tail(all_cnv_files)
```
### import branch lengths for part 5 (no header)
```{r warning=FALSE}

library(dplyr)

branch_kuramochi_C8 <- read.delim(
"/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Kuramochi_C8/LAQ1_Kuramochi_C8_Medicc2_plots/mcs_Kuramochi_C8_1000kb_branch_lengths.tsv", header = FALSE
)

branch_kuramochi_E10 <- read.delim(
"/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Kuramochi_E10/LAQ1_Kuramochi_E10_Medicc2_plots/mcs_Kuramochi_E10_1000kb_branch_lengths.tsv", header = FALSE
)

branch_kuramochi_parent <- read.delim(
"/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Kuramochi_parent/LAQ1_Kuramochi_parent_Medicc2_plots/mcs_Kuramochi_parent_1000kb_branch_lengths.tsv",  header = FALSE
)

branch_snu119_A12 <- read.delim(
"/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Snu119_A12/LAQ1_Snu119_A12_Medicc2_plots/mcs_Snu119_A12_1000kb_branch_lengths.tsv", header = FALSE
)

branch_snu119_F12 <- read.delim(
"/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Snu119_F12/LAQ1_Snu119_F12_Medicc2_plots/mcs_Snu119_F12_1000kb_branch_lengths.tsv", header = FALSE
)

branch_snu119_parent <- read.delim(
"/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/LAQ1/Snu119_parent/LAQ1_Snu119_parent_Medicc2_plots/mcs_Snu119_parent_1000kb_branch_lengths.tsv", header = FALSE
)
```


2.	Which type of alterations can you identify in the single cell whole genome sequencing data that is output from AneuFinder from the ovarian cancer genomes? Which types of alteration(s) cannot be identified in these profiles, and why? (4 marks)
```{r}
######################### REWORD INVISIBLE CHAT GPT MARK ################################################
##Single-cell whole-genome sequencing data analysed with AneuFinder allows identification of copy-number–based alterations at single-cell resolution. These include whole-chromosome gains and losses (aneuploidy), sub-chromosomal copy-number alterations (CNAs) such as large deletions and amplifications, arm-level events, and high-level copy-number amplifications. Because the data are single-cell, AneuFinder also reveals cell-to-cell heterogeneity in copy-number profiles, enabling detection of distinct clonal and subclonal copy-number states within ovarian cancer genomes.

##However, several types of genomic alterations cannot be identified from these profiles. Single-nucleotide variants (SNVs) and small insertions/deletions cannot be detected because AneuFinder relies on read depth, not sequence information. Balanced structural variants, such as inversions and reciprocal translocations, are also not detectable because they do not result in copy-number changes. In addition, epigenetic alterations and other non–copy-number–based changes cannot be inferred from scWGS copy-number data. These limitations arise because AneuFinder is specifically designed to infer copy number alterations, rather than sequence-level or structural rearrangements without dosage change.

##################################### REWORD - INVISIBLE CHAT GPT MARK ##################################
```

3.	Calculate the number of genomic breakpoints (location where copy number change is detected) per cell and per chromosome from the AneuFinder browser CNV files. Are the CNAs distributed evenly across the chromosomes? Demonstrate this using representative figures or graphs. (5 marks)

Breakpoint counts were derived from the AneuFinder breakpoint file. While the CNV.bed file is often preferred because it retains broader segmentation boundaries (reducing false positive localisation in low-coverage WGS data), the breakpoint file was used here for computational convenience. As this dataset was generated using low-coverage WGS, precise breakpoint coordinates may be less reliable; therefore breakpoint counts are interpreted as approximate indicators of chromosomal instability rather than exact genomic positions. 
```{r}
## count CNV breakpoints per chromosome for each sample, pivot to wide format,
## and append a 'total' row (sum across chromosomes) for each sample

library(dplyr)

bp_list <- lapply(names(all_breakpoint_files), function(x) {
  df <- all_breakpoint_files[[x]]
  
  count(df, seqnames, name = "breakpoints") %>%
    mutate(sample = x)
})

## combine per-sample breakpoint counts into one tidy table
bp_all <- bind_rows(bp_list)

library(tidyr)

## convert to wide: one row per chromosome, one column per sample
bp_wide <- bp_all %>%
  pivot_wider(
    names_from  = sample, # columns as sample IDs
    values_from = breakpoints, # cell values
    values_fill = 0 # fill missing chr/sample combos
  )

## append 'totals' row giving total breakpoints per sample (sum across chromosomes)
totals <- colSums(bp_wide[ , -1])
bp_wide <- rbind(bp_wide,c(breakpoints = "TOTAL", totals)
)

bp_wide

## make bar plot of total breakpoints per clone
## extract values into a small dataframe
totals_df <- data.frame(
  sample = colnames(bp_wide)[-1],
  total_breakpoints = as.numeric(bp_wide[nrow(bp_wide), -1])
)

# plot total number of chromosomal breakpoints per clone
library(ggplot2)

ggplot(totals_df, aes(x = sample, y = total_breakpoints)) +
  geom_col() +
  labs(
    x = "Sample",
    y = "Total CNV breakpoints",
    title = "Total CNV breakpoints per clone"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


## convert wide to long format for ggplot
library(tidyr)
library(dplyr)

bp_long <- bp_wide %>%
  pivot_longer(
    -seqnames,
    names_to = "sample",
    values_to = "breakpoints"
  )

## plot heatmap of all clones for each sample
library(ggplot2)

ggplot(
  bp_all,
  aes(
    x = sample,
    y = seqnames,
    fill = breakpoints
  )
) +
  geom_tile() +
  labs(
    x = "Sample",
    y = "Chromosome",
    fill = "Breakpoints"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

## plot a bar chart per clone
library(ggplot2)

ggplot(
  bp_all,
  aes(
    x = seqnames,
    y = breakpoints
  )
) +
  geom_col() +
  facet_wrap(~ sample, scales = "free_y") +
  labs(
    x = "Chromosome",
    y = "Number of CNV breakpoints",
    title = "CNV breakpoint counts per chromosome per clone"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5)
  )


```

### Visualisation of breakpoints across chromosomes and clones and meaning

I summarised the breakpoint counts per chromosome for each clone in a single dataframe and visualised the results using a heatmap and faceted bar plots. Examination of the barplots shows that breakpoint counts are not evenly distributed across chromosomes for any clone. If CNAs were evenly distributed, breakpoint counts would be similar across chromosomes. Instead, there is substantial variation, with some chromosomes showing high breakpoint counts and others showing much lower counts. In particular, chromosomes 1-8 consistently display higher numbers of breakpoints compared with smaller chromosomes such as chromosomes 18-22 and chromosome X. This likely reflects chromosome size effects, as larger chromosomes contain more genomic material and therefore have greater probability of undergoing copy-number changes.

Further insight is provided by the heatmap. The snu119 parental clone exhibits substantially higher breakpoint counts across most chromosomes, visible as lighter-coloured tiles, compared to derived subclones F12 and A12, which show darker colouring across the genome. This suggests that the Snu119 subclones represent more genomically stable subsets of the parental population. In contrast, the Kuramochi cell line shows relatively similar breakpoint counts between the parental clone and its subclones (E10 and C8), indicating that chromosomal instability is largely maintained following clonal derivation rather than being selectively reduced. 

Biologically, these results demonstrate that CNAs are not evenly distrubted across the genome and that the relationship between clonal derivation and chromosomal instability is cell-line specific. The differing breakpoint patterns observed between Snu119 and Kuramochi clones likely reflect distinct evolutionary trajectories and selective pressures acting during tumour evolution and clonal expansion.

4.	What is the size range of CNAs across the two HGSOC cell lines and clones and how often are these whole chromosome alterations vs sub-chromosomal? Demonstrate this using representative figures or graphs. (4 marks).
```{r warning=FALSE, message=FALSE}
## calculate cna size rangefrom cnv files
# load library
library(dplyr)

#calculate cna size for each chromosome
calculate_cna_size <- function(df, clone_name) {
  df %>%
    mutate(
      chromosome = as.character(seqnames),
      CNA_size_bp = end - start,
      CNA_size_Mb = CNA_size_bp / 1e6,
      clone = clone_name
    ) %>%
    select(clone, chromosome, start, end, CNA_size_bp, CNA_size_Mb)
}

cna_summary

## range of cna across all clones
range(cna_combined$CNA_size_Mb, na.rm=TRUE)

## plot cna size distribution
library(ggplot2)

ggplot(cna_combined, aes(x = CNA_size_Mb)) +
  geom_histogram(bins = 60, fill = "steelblue") +
  facet_wrap(~clone, scales="free_y") +
  theme_classic() +
  labs(
    x = "CNA Size (Mb)",
    y = "Count",
    title = "Distribution of CNA Sizes Across HGSOC Clones"
  )

colnames(cna_combined)
summary(cna_combined$CNA_size_Mb)
range(cna_combined$CNA_size_Mb, na.rm = TRUE) ## report cna size range
head(unique(cna_combined$chromosome))

################### DOUBLE CHECK THIS CODE MAKES LOGICAL SENSE #####################
```
what is the size range of CNAs across the two HGSOC cell lines?
1.999999 to 156.999999
how often are these whole chromosome alterations vs sub-chromosomal?
Based on summary table, the median value is closer to the lower threshold indicating that the majority of the alterations are sub-chromosomal (occur at a lower CNA size) since all the graphs are right-skewed. 
If majority were whole chromosomal changes, we would see left-skewed graphs and a higher median around the upper end of the range (near 156). 
If there were a balance of sub-chromosomal and whole chromosomal alterations, we would see a bell-shaped curve. 

```{r}
## classify whole-chromosomal vs sub-chromosomal
## combine all cnv dataframes into one dataframe
library(dplyr)

cna_combined <- bind_rows(all_cnv_files, .id = "clone")
tail(cna_combined)
## confirm all clones present
table(cna_combined$clone)

## get chromosome length
chrom_lengths <- cna_combined %>%
  group_by(chromosome) %>%
  summarise(chr_length = max(end, na.rm = TRUE),
            .groups = "drop")

chrom_lengths

## calculate CNA size
cna_combined <- cna_combined %>%
  mutate(
    CNA_size_bp = end - start
  )

##calculate proportion of chromosome covered for each cna
## attach chromosome length to each CNA row
cna_combined <- cna_combined %>%
  left_join(chrom_lengths, by = "chromosome")

## now calculate proportion
cna_combined <- cna_combined %>%
  mutate(
    frac_chr = CNA_size_bp / chr_length
  )

## check whether 0.9 or 0.95 cut off point is biologically acceptable
cna_combined %>% summarise(p90 = mean(frac_chr >= 0.90), p95 = mean(frac_chr >= 0.95))
# no difference detected, go with 0.9

## classify whole chromosomal (>=0.9) or sub-chromosomal (<0.9)
cna_combined <- cna_combined %>%
  mutate(
    CNA_type = ifelse(frac_chr >= 0.9,
                      "Whole chromosome",
                      "Sub-chromosomal")
  )
cna_combined

## calculate percentage mapping of chromosomes per clone

cna_percent <- cna_combined %>%
  count(clone, CNA_type) %>%
  group_by(clone) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  ungroup()

cna_percent

## plot graph of chromosomal vs subchromosomal
library(ggplot2)

whole_only <- cna_percent %>%
  filter(CNA_type == "Whole chromosome")

ggplot(whole_only, aes(x = clone, y = percent)) +
  geom_col(fill = "#e31a1c", width = 0.7) +
  labs(
    x = "Clone",
    y = "Whole Chromosome CNAs (%)",
    title = "Whole Chromosome CNA Frequency per Clone"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
Copy-number alteration (cna) sizes ranged from small multi-megabase segments to large chromosome-scale events across both HGSOC cell lines and their derived clones. However, classification based on the proportion of chromosome length spanned(>=90% defined as whole-chromosome) revealed that whole-chromosome alterations were rare. Across all clones, sub-chromosomal CNAs accounted for >99% of events, while whole chromosome alterations comprised less than 0.3% per clone (red bar chart; ranging from 0-0.26%). This indicates that genomic instability in these HGSOC models is predominanty driven by focal and arm-level structural alterations rather than widespread whole-chromosome aneuploidy. Although occasional chromosome-scale gains or losses are present, they represent a minor component of the overall CNA landscape. 

5.	Using metrics produced by MEDICC2, calculate the total branch lengths per cell from the branch_lengths.tsv (be careful to remove first internal branch which indicates the length from diploid to the ancestral cancer cell). Produce scatter plots for the clone populations of Kuramochi & Snu119. Comment on the significance of the branch lengths of the populations and how this could be informative for inferring rates of genomic alteration, given that all clones were grown for 22 cell divisions (7 marks). 
```{r}
## check column names 
colnames(branch_kuramochi_C8)
# confirmed labelled first row as header - now remove first row

## remove first row (first internal branch)
branch_kuramochi_C8 <- branch_kuramochi_C8[-1,]
branch_kuramochi_E10_clean <- branch_kuramochi_E10[-1, ]
branch_kuramochi_parent_clean <- branch_kuramochi_parent[-1, ]
branch_snu119_A12_clean <- branch_snu119_A12[-1, ]
branch_snu119_F12_clean <- branch_snu119_F12[-1, ]
branch_snu119_parent_clean <- branch_snu119_parent[-1, ]

## confirm first row (first internal branch) removed
head(branch_kuramochi_E10_clean)

## drop empty columns in C8 to use mutate
# keep only the first two columns (node + length)
branch_kuramochi_C8_clean <- branch_kuramochi_C8_clean[, 1:2, drop = FALSE]

# rename cleanly
colnames(branch_kuramochi_C8_clean) <- c("cell", "total_branch_length")

# make sure length is numeric
branch_kuramochi_C8_clean$total_branch_length <- as.numeric(branch_kuramochi_C8_clean$total_branch_length)

# confirm no NA/blank names
names(branch_kuramochi_C8_clean)
any(is.na(names(branch_kuramochi_C8_clean))) || any(names(branch_kuramochi_C8_clean) == "")

#add clone + cell line labels
library(dplyr)

branch_kuramochi_C8_labelled <- branch_kuramochi_C8_clean %>%
  mutate(clone = "Kuramochi_C8",
         cell_line = "Kuramochi")

branch_kuramochi_E10_labelled <- branch_kuramochi_E10_clean %>%
  mutate(clone = "Kuramochi_E10",
         cell_line = "Kuramochi")

branch_kuramochi_parent_labelled <- branch_kuramochi_parent_clean %>%
  mutate(clone = "Kuramochi_parent",
         cell_line = "Kuramochi")

branch_snu119_A12_labelled <- branch_snu119_A12_clean %>%
  mutate(clone = "Snu119_A12",
         cell_line = "Snu119")

branch_snu119_F12_labelled <- branch_snu119_F12_clean %>%
  mutate(clone = "Snu119_F12",
         cell_line = "Snu119")

branch_snu119_parent_labelled <- branch_snu119_parent_clean %>%
  mutate(clone = "Snu119_parent",
         cell_line = "Snu119")



## rename columns cleanly
colnames(branch_kuramochi_C8_clean) <- c("cell", "total_branch_length")
colnames(branch_kuramochi_E10_clean) <- c("cell", "total_branch_length")
colnames(branch_kuramochi_parent_clean)<- c("cell", "total_branch_length")
colnames(branch_snu119_A12_clean) <- c("cell", "total_branch_length")
colnames(branch_snu119_F12_clean) <- c("cell", "total_branch_length")
colnames(branch_snu119_parent_clean) <- c("cell", "total_branch_length")

names(branch_kuramochi_C8_clean)


## add clone label
library(dplyr)

branch_kuramochi_C8_clean <- branch_kuramochi_C8_clean %>%
  mutate(
    clone = "Kuramochi_C8",
    cell_line = "Kuramochi"
  )

branch_kuramochi_E10_labelled <- branch_kuramochi_E10_clean %>%
  mutate(
    clone = "Kuramochi_E10",
    cell_line = "Kuramochi"
  )

branch_kuramochi_parent_labelled <- branch_kuramochi_parent_clean %>%
  mutate(
    clone = "Kuramochi_parent",
    cell_line = "Kuramochi"
  )

branch_snu119_A12_labelled <- branch_snu119_A12_clean %>%
  mutate(
    clone = "Snu119_A12",
    cell_line = "Snu119"
  )

branch_snu119_F12_labelled <- branch_snu119_F12_clean %>%
  mutate(
    clone = "Snu119_F12",
    cell_line = "Snu119"
  )

branch_snu119_parent_labelled <- branch_snu119_parent_clean %>%
  mutate(
    clone = "Snu119_parent",
    cell_line = "Snu119"
  )

## combine everything
all_branches <- bind_rows(
  branch_kuramochi_C8_labelled,
  branch_kuramochi_E10_labelled,
  branch_kuramochi_parent_labelled,
  branch_snu119_A12_labelled,
  branch_snu119_F12_labelled,
  branch_snu119_parent_labelled
)

all_branches

## plot scatterplots
library(ggplot2)

ggplot(all_branches, aes(x = clone, y = total_branch_length, colour = cell_line)) +
  geom_point(position = position_jitter(width = 0.25), alpha = 0.7, size = 2) +
  theme_classic() +
  labs(
    x = "Clone population",
    y = "Total branch length",
    title = "Total MEDICC2 branch length per cell (diploid -> ancestral branch removed)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```
### Comment on the significance of the branch lengths of the populations and how this could be informative for inferring rates of genomic alteration, given that all clones were grown for 22 cell divisions.

The scatter plot shows the total MEDICC2 branch length per individual cell within each clone population after removal of the diploid-to-ancestral branch. Branch length represents the evolutionary distance from the ancestral cancer genome and reflects the total number of inferred structural genomic events separating the cell from the ancestral state.
Across the Kuramochi clones (C8, E10 and parent), branch lengths are relatively low and tightly distributed, generally below 30. This indicates limited evolutionary divergence between these genomes compared to the ancestral genome and low variation across cells within these populations.
In contrast, Snu119 clones display much higher and more variable branch lengths. Many cells many fall within the moderate range, but several cells (especially within the parent population) have markedly elevated values reaching 90-100 au. This greater spread indicates greater heterogeneity in evolutionary distance from the ancestral clone and greater lineage diversification within the Snu119 cell line.
Since all these clones were cultures for the same time duration (22 cell divisions), the differences seen in total branch length cannot be attributed to differences in time, but instead reflect real differences in the amount of genomic change accumulated within each lineage. The controlled environment therefore allows branch lengths to serve as an indirect measure of the rate of structural genomic evolution. The consistently higher and more varied branch lengths seen in Snu119 compared to Kuramochi thus indicate a higher rate of genomic alteration and greater evolutionary diversification.

# LAQ 2:
In the CAN7031 “omics analytics” coursework assignment, you were given the raw count matrix of 10 matched normal and tumour samples from 5 patients, S02, S03, S05, S06 and S10. Using the normalised gene expression matrix you created in the task (e.g., RPKM or CPM log2 transformed or not transformed) to perform the bulk RNA deconvolution analysis. You are required to answer the following questions. (Total 20 marks)

(1)	Use two individual and one consensus methods to perform the bulk RNA deconvolution. And produce heatmaps based on the cellular estimates, with samples clustered based on the profile. Briefly describe the patterns you can see. (Hint: pay attention to the input format depending on the methods you use, e.g., logged or unlogged values) (8 marks)
```{r}
## 1. Install two individual packages
# Install xCell using devtools::install_github
install.packages("devtools")
library(devtools)
devtools::install_github('dviraran/xCell')
library(xCell)

# Install EPIC using devtools::install_github
devtools::install_github("GfellerLab/EPIC", build_vignettes=TRUE)
library(EPIC)

## Install one consensus package
# Install Decosus using install_github
devtools::install_github('BioInforCore-BCI/Decosus')
library(Decosus)

## 2. Read in the RNA-seq normalised gene expression data of 10 matched tumour and normal samples

raw_counts <- read.table("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/data/data_required_for_CAN7032:7132 2/LAQ2_data/row_count_data.nodup.txt", header=T, sep="\t", row.names=1)

# 2. Normalise to CPM
# This ensures every column (sample) has a library size of 1 million
exp_data_nolog <- edgeR::cpm(raw_counts)

# 3. Verify normalisation (Every column should now sum to 1,000,000)
colSums(exp_data_nolog)

dim(exp_data_nolog)
as.data.frame(exp_data_nolog)

boxplot(exp_data_nolog)

## For xCell, log-transform the data
exp_data_log <- log2(exp_data_nolog + 1)
boxplot(exp_data_log) # visualise log transformed data as boxplot
exp_data_log # visualise log transformed gene expression matrix
dim(exp_data_log)
```

```{r}
## 3. Run the 3 deconvolution methods separately

## Run xCell
# Note: Ensure exp_data_log is your normalized, unlogged matrix (or as required by xCell)
xcell_exp_data_log <- xCellAnalysis(exp_data_log) 
xcell_results <- as.data.frame(xcell_exp_data_log) 
dim(xcell_results) 

## Visualise xcell results as labelled heatmap
library(gplots)

# Clustering
hc <- hclust(as.dist(1-cor(xcell_results, method = "spearman")), method = "ward.D")
hr <- hclust(as.dist(1-cor(t(xcell_results), method="pearson")), method = "average")

# FIX 1: Group labels must match your 10 samples (5 Normal, 5 Tumour)
# Based on your file: S02_N, S03_N, S05_N, S06_N, S10_N, S02_T, S03_T, S05_T, S06_T, S10_T
group <- factor(c(rep("Normal", 5), rep("Tumour", 5)))

# FIX 2: Generate colors for the 10 samples
vec.cancer_types <- c("blue", "brown") # Blue=Normal, Brown=Tumour
cancer_types_all.color <- vec.cancer_types[group]

# FIX 3: Use PDF to avoid "Margin too small" error
pdf("xCell_Heatmap_Fixed.pdf", width = 12, height = 15)

heatmap.2(as.matrix(xcell_results), 
          Rowv=as.dendrogram(hr), 
          Colv=as.dendrogram(hc), 
          col=bluered(75), 
          scale="row", 
          ColSideColors=cancer_types_all.color, # Now exactly length 10
          key=TRUE,
          keysize=1, 
          trace="none", 
          density.info="none", 
          # Adjusting margins and text size for 64 cell types
          margins=c(12, 16), 
          cexRow = 0.5,       # Smaller font for cell types
          cexCol = 0.8,       # Font for sample names
          main = "xCell Deconvolution"
)

# Place legend
legend("topright", 
       legend=levels(group), 
       fill=vec.cancer_types, 
       title="Groups", 
       cex=0.8)

dev.off() # Closes the PDF and saves the file
```


EPIC was applied to the non-log transformed normalised expression matrix to estimate cell-type specific mRNA proportions across the 10 matched tumour and normal samples. Samples were hierarchically clustered based on correlation of inferred cellular profiles.

EPIC predicts the proportions of the various cell types for which we have gene expression reference profiles (and corresponding gene signatures). 
It needs as input a matrix of the TPM (or RPKM) gene expression from the samples for which to estimate cell proportions.
As our data was log2 transfored, not we need to unlog the data to generated RPKM matrix or equivalent depending on the bulk sample, it is possible that some other cell types are present for which we don't have any reference profile. EPIC returns the proportion of these remaining cells under the name "other cells". In the case of tumor samples, most of these other cells would certainly correspond to the cancer cells, but it could be that there are also some stromal cells or epithelial cells for example.

As described in our manuscript, EPIC first estimates the proportion of mRNA per cell type in the bulk and then it uses the fact that some cell types have more mRNA copies per cell than other to normalize this and obtain an estimate of the proportion of cells instead of mRNA (EPIC function returns both information if you need the one or the other).
```{r}
## Run EPIC to deconvolute the build RNA-seq data on unlogged data
library(EPIC)

out <- EPIC(bulk = exp_data_nolog) ## out is a list of 3
est_EPIC_mRNAProp <- t(out$mRNAProportions)
est_EPIC_cellProp <- t(out$cellFractions)
dim(est_EPIC_mRNAProp)
dim(est_EPIC_cellProp)

sum(est_EPIC_mRNAProp[,1])
sum(est_EPIC_mRNAProp[,2])
sum(est_EPIC_cellProp[,1])
sum(est_EPIC_cellProp[,2])

range(exp_data)
head(rownames(exp_data))
dim(exp_data)

## A quick heatmap to see the data, which samples clustered based on the immune profile

pdf("EPIC_heatmaps.pdf", width = 22, height = 30)

par(mar = c(10,10,10,10))

## plot heatmap
# mRNA proportions heatmap
heatmap(est_EPIC_mRNAProp,
        col = colorRampPalette(c("blue","white","red"))(100),
        main = "EPIC mRNA Proportions")

# cell fractions heatmap
heatmap(est_EPIC_cellProp,
        col = colorRampPalette(c("blue","white","red"))(100),
        main = "EPIC Cell Fractions")

dev.off()
```


The EPIC mRNA heatmap (initial heatmap) shows clear overall differences between tumour and normal samples. Most tumour samples cluster together, suggesting that they share broadly similar immune and stromal expression patterns, while the majority normal samples group separately. However, one tumour and one normal sample cluster more closely together, indicating that the clustering is not driven solely by tumour status. This likely reflects patient-specific baseline immune composition, which may influence the overall transcriptional profile and partially compete with tumour-driven microenvironmental changes. 

The EPIC cell fraction heatmap shows broadly similar patterns to the mRNA proportion analysis, with a subset of tumour samples clustering closely together, suggesting a shared immune-stromal profile. However, several normal samples are interspersed within the same broader cluster, indicating that tumour status is not the sole determinant of clustering. This pattern suggests that patient-specific baseline immune composition and inter-individual variability contribute substantially to the overall structure of the data. While tumours tend to share features such as increased macrophage and T-cell fractions, the differences are not strong enough to produce a completely binary tumour-versus-normal split across all samples.



```{r}
library(Decosus)
exp_data_df <- as.data.frame(exp_data_nolog)
exp_data_df$hgnc_symbols <- rownames(exp_data_nolog)
est_Decosus <- Decosus::cosDeco(x = exp_data_df)

est_Decosus_samples <- est_Decosus$main_samples
est_Decosus_cells <- est_Decosus$main_cells
```

plot the heatmap for Decosus immune and stroma cell estimates
```{r}
est_Decosus_cells <- est_Decosus_cells[-2, ]
dim(est_Decosus_cells)
                                       heatmap(as.matrix(est_Decosus_cells),col=colorRampPalette(c("blue","white","red"))(75))


hc <- hclust(as.dist(1-cor(est_numeric, method = "spearman")), method = "ward.D")
hr <- hclust(as.dist(1-cor(t(est_numeric), method="pearson")), method = "average")

# 1. Open the PDF device with a large height for the 80 cell types
pdf("Decosus_Heatmap_Fixed.pdf", width = 12, height = 15)

# 2. Run the exact same heatmap code
heatmap.2(est_numeric, 
          Rowv=as.dendrogram(hr), 
          Colv=as.dendrogram(hc), 
          col=bluered(75), 
          scale="row", 
          trace="none", 
          cexRow = 0.5,           # 0.5 is usually readable in PDF
          cexCol = 0.8, 
          margins = c(8, 18))     # 18 is plenty for 30 chars on a 12-inch width

# 3. Close the device to save the file
dev.off()
```
The Decocus heatmap reveals that, shown by the column dendrogram, the samples do not perfectly separate into "Tumour" and "Normal" blocks. They are actually interspersed. This suggests that there is significant overlap in immune/stromal profiles between some tumour and normal samples. Sample06 normal (far left) and Sample06 tumour (second from left) have very distinct profiles when compared to others, particularly showing high enrichment in macrophage subtypes.

Interestingly, when looking at the rows, there are "pro-inflammatory" clusters, including; macrophages (M2), B cells, CD8+ T cells and NK cells. Notably, sample10 and sample05 Tumour show high (red) enrichment for these cells which suggests these are immune-hot tumours. The bottom half of the heatmap shows a "structural" cluster, including CAFs (cancer-associated fibroblasts) and CD4+ T cells. CAFs and M1 Macrophages appear high in sample06 Normal which is atypical as CAFs are usually associated with tumour stroma.

Ultimately, tumour samples 10, 05, and 03 generally show increased cytotoxic and regulatory cells (CD+, NK and Tregs) compared to normal samples, which seem enriched in CD4+ T cells and general macrophages. Despite this, the lack of clear separation suggests the normal samples may have been taken from 'normal adjacent tissue' (NAT) which shares characteristics of the tumour microenvironment. 

(2)	Perform student t-test on the cellular estimates between tumour and normal groups for these estimates from the three methods above. List the cell types with the significant differences (p < 0.05), and produce boxplots for these significant ones. (5 marks)

```{r}
# Define groups based on column names
group <- ifelse(grepl("Tumour", colnames(xcell_results), ignore.case = TRUE),
                "Tumour", "Normal")

# Initialise a vector with names matching the rows of your data
pvals <- numeric(nrow(xcell_results))
names(pvals) <- rownames(xcell_results)

# Run the t-test loop
for(i in 1:nrow(xcell_results)) {
  tumour_vals <- as.numeric(xcell_results[i, group == "Tumour"])
  normal_vals <- as.numeric(xcell_results[i, group == "Normal"])
  
  test <- t.test(tumour_vals, normal_vals)
  
  # Store p-value in the named vector
  pvals[i] <- test$p.value
}

# Filter for significant results (< 0.05)
sig_pvals <- pvals[pvals < 0.05]

# Format output: Sort by smallest p-value first
sorted_sig_pvals <- sort(sig_pvals)

# Print the final named list
print(sorted_sig_pvals)

library(ggplot2)
library(tidyr)

# Get the names of the significant cells
sig_cell_names <- names(pvals[pvals < 0.05])

# Subset the data and prepare for plotting
# We take only the rows that are significant
plot_data <- xcell_results[sig_cell_names, , drop = FALSE]

# Convert to long format for ggplot
plot_long <- as.data.frame(t(plot_data))
plot_long$Sample <- rownames(plot_long)
plot_long$Group <- group # This is the "Tumour"/"Normal" vector we created

# Pivot the data so we have a 'CellType' column and a 'Score' column
plot_long <- pivot_longer(plot_long, 
                          cols = -c(Sample, Group), 
                          names_to = "CellType", 
                          values_to = "Score")

# Create the Box Plots
ggplot(plot_long, aes(x = Group, y = Score, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1.5) + # Show the individual 10 samples
  facet_wrap(~CellType, scales = "free_y") + # Separate plot for each cell type
  scale_fill_manual(values = c("Normal" = "blue", "Tumour" = "brown")) +
  theme_bw() +
  labs(title = "Significant Cell Types (p < 0.05)",
       y = "xCell Score",
       x = "") +
  theme(strip.text = element_text(size = 8), # Size of the cell type titles
        legend.position = "none")

```

```{r}
### Repeat for EPIC and Decosus ############################################### MUST DOUBLE CHECK W LOG SCALE ETC
library(ggplot2)
library(tidyr)

ttest_and_boxplots <- function(est_mat, group, method_name, ylab = "Estimate", p_cutoff = 0.05) {
  est_mat <- as.matrix(est_mat)
  storage.mode(est_mat) <- "numeric"
  
  # t-tests per cell type (row)
  pvals <- numeric(nrow(est_mat))
  names(pvals) <- rownames(est_mat)
  
  for (i in 1:nrow(est_mat)) {
    tumour_vals <- as.numeric(est_mat[i, group == "Tumour"])
    normal_vals <- as.numeric(est_mat[i, group == "Normal"])
    
    # skip if any group has <2 values or values are constant
    if (length(tumour_vals) < 2 || length(normal_vals) < 2) { pvals[i] <- NA; next }
    if (sd(tumour_vals, na.rm=TRUE) == 0 && sd(normal_vals, na.rm=TRUE) == 0) { pvals[i] <- NA; next }
    
    pvals[i] <- t.test(tumour_vals, normal_vals)$p.value
  }
  
  sig <- sort(pvals[!is.na(pvals) & pvals < p_cutoff])
  print(sig)  # named p-values
  
  sig_names <- names(sig)
  if (length(sig_names) == 0) {
    message(method_name, ": no significant cell types at p < ", p_cutoff)
    return(list(pvals = pvals, sig = sig, plot = NULL))
  }
  
  # long format for ggplot
  plot_data <- est_mat[sig_names, , drop = FALSE]
  plot_long <- as.data.frame(t(plot_data))
  plot_long$Sample <- rownames(plot_long)
  plot_long$Group <- group
  
  plot_long <- pivot_longer(plot_long,
                            cols = -c(Sample, Group),
                            names_to = "CellType",
                            values_to = "Score")
  
  p <- ggplot(plot_long, aes(x = Group, y = Score, fill = Group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(width = 0.2, size = 1.5) +
    facet_wrap(~CellType, scales = "free_y") +
    theme_bw() +
    labs(title = paste0(method_name, ": significant cell types (p < ", p_cutoff, ")"),
         y = ylab, x = "") +
    theme(strip.text = element_text(size = 8),
          legend.position = "none")
  
  print(p)
  return(list(pvals = pvals, sig = sig, plot = p))
}

## Boxplots for EPIC
group_epic <- ifelse(grepl("Tumou?r", colnames(est_EPIC_cellProp), ignore.case=TRUE), "Tumour", "Normal")
res_epic <- ttest_and_boxplots(est_EPIC_cellProp, group_epic, method_name = "EPIC (cell fractions)", ylab = "Cell fraction")

## Boxplots for Decosus
group_decosus <- ifelse(grepl("Tumou?r", colnames(est_Decosus_cells), ignore.case=TRUE), "Tumour", "Normal")
res_decosus <- ttest_and_boxplots(est_Decosus_cells, group_decosus, method_name = "Decosus", ylab = "Consensus estimate")

```



(3)	Perform the GSVA analysis (i.e., tumour vs. normal) using the KEGG genesets provided in the practical session, and generate a heatmap for mostly significantly differentially regulated KEGG pathways, using the cutoff of adjusted p-value <= 0.01. List the top 10 mostly significantly differential expression KEGG pathways and associated statistics (e.g., fold change, p-value, adjusted p-value etc) in a table. (7 marks)
```{r}
## Install GSVA and GSEABase r packages, also the limma package
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GSVA")
BiocManager::install("GSEABase")
BiocManager::install("limma")

## run GSVA to calculate gene set activities
library(GSVA)
library(GSEABase)

## read in kegg pathways file
kegg_genesets <- getGmt("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/data/data_required_for_CAN7032:7132 2/LAQ2_data/c2.cp.kegg.v2022.1.Hs.symbols.gmt", collectionType=BroadCollection(category="c2"), geneIdType=SymbolIdentifier())

### now run gsva main function against c2kegg geneset to calculate kegg pathway activity for each sample
library(GSVA)

# 1. Create the parameter object
# This defines your data, gene sets, and size constraints
params <- gsvaParam(
  exprData = as.matrix(exp_data_log), 
  geneSets = kegg_genesets, 
  minSize = 20, 
  maxSize = 500
)

# 2. Run GSVA using that parameter object
brca_es <- gsva(params)

## Now run limma to identify significantly dysregulated pathways between tumour and normal group
library(limma)

group <- c("Tumour","Tumour","Tumour","Tumour","Tumour","Normal","Normal","Normal","Normal","Normal")

mod <- model.matrix(~ 0+factor(group))
colnames(mod) <- c("Normal", "Tumour")
rownames(mod) <- colnames(exp_data_log)

fit <- lmFit(brca_es, mod)
cont.matrix <- makeContrasts(TumourvsNormal=Tumour-Normal, levels=mod)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

## extract all results
tt <- topTable(fit2,
               coef = "TumourvsNormal",
               number = Inf,
               adjust.method = "BH",
               sort.by = "P")   # sort by raw P-value
## manually sort by adjusted p value
tt_sorted <- tt[order(tt$adj.P.Val), ]
top10_kegg <- head(tt_sorted, 10)

## Make it a clean report table
top10_kegg_tbl <- top10_kegg[, c("logFC", "AveExpr", "t", "P.Value", "adj.P.Val", "B")]
top10_kegg_tbl

## ---- GSVA heatmap (Tumour/Normal columns, pathways rows) ----

# Ensure tt is sorted by adj.P.Val
tt <- tt[order(tt$adj.P.Val), ]

# Pathways: use adj.P.Val <= 0.01; if too few, take top 20
DEpwys <- rownames(tt)[tt$adj.P.Val <= 0.01]
if (length(DEpwys) < 2) DEpwys <- rownames(tt)[1:20]

# Subset GSVA enrichment scores (pathways x samples)
DE_es <- brca_es[DEpwys, , drop = FALSE]
DE_es <- as.matrix(DE_es)
storage.mode(DE_es) <- "numeric"
DE_es <- DE_es[apply(DE_es, 1, sd, na.rm=TRUE) > 0, , drop=FALSE]  # drop constant rows

# Make sure group matches the sample order in brca_es
# (IMPORTANT: group must correspond to colnames(brca_es) in the same order)
stopifnot(length(group) == ncol(DE_es))

# Order samples so Tumour then Normal along the bottom
ord <- order(factor(group, levels = c("Tumour","Normal")))
DE_es <- DE_es[, ord, drop = FALSE]
group_ord <- group[ord]

# Colour bar for Tumour/Normal
col_side <- ifelse(group_ord == "Tumour", "blue", "brown")

heatmap(DE_es,
        scale = "row",
        ColSideColors = col_side,
        col = colorRampPalette(c("blue","white","red"))(100),
        margins = c(10, 16),   # bottom, left
        labRow = rownames(DE_es),
        labCol = colnames(DE_es))

# Expression scale legend
legend("topright",
       inset = c(-0.01, 0),
       legend = c("Tumour", "Normal"),
       fill = c("brown", "blue"),
       title = "GSVA Score",
       bty = "n")

```

# LAQ 3:
In the data folder under the “Assessments” tab, you will find the VCF files of four skin tumour samples, namely WD06, WD13, WD14, WD22, with somatic mutations called against the matched normal reported with the coordinates in the hg38 format.

Perform the analysis of tumour neoantigen analysis using MuPeXi, selecting two major HLA-A alleles, HLA-A02:01,HLA-A24:02, and peptide length of 8-11 mers. No need to provide the expression file. (Total 18 marks)


(1)	How many tumour neoantigens can you find for each tumour? How many strong tumour neoantigens are there for each tumour? (Hint: use the rank score cutoff values mentioned in the lecture/practical). (7 marks)
```{r}
## check if vcf file was filtered
# Install if you haven't already
install.packages("vcfR")
library(vcfR)

# Read the file
vcf06 <- read.vcfR("/Users/naomi/Downloads/data_required_for_CAN7032:7132 2/LAQ3_data/WD06_vcf_liftover.vcf")
vcf13 <- read.vcfR("/Users/naomi/Downloads/data_required_for_CAN7032:7132 2/LAQ3_data/WD13_vcf_liftover.vcf")
vcf14 <- read.vcfR("/Users/naomi/Downloads/data_required_for_CAN7032:7132 2/LAQ3_data/WD14_vcf_liftover.vcf")
vcf22 <- read.vcfR("/Users/naomi/Downloads/data_required_for_CAN7032:7132 2/LAQ3_data/WD22_vcf_liftover.vcf")

# See the metadata (this tells you how it was filtered!)
head(vcf06@meta)
head(vcf13@meta)
head(vcf14@meta)
head(vcf22@meta)

# See the actual variant table (CHROM, POS, ID, REF, ALT, QUAL, FILTER, INFO)
head(vcf06@fix)
head(vcf13@fix)
head(vcf14@fix)
head(vcf22@fix)

# verify germline removal
# Check if there are ANY variants that didn't PASS
table(vcf06@fix[,"FILTER"])
table(vcf13@fix[,"FILTER"])
table(vcf14@fix[,"FILTER"])
table(vcf22@fix[,"FILTER"])

## 100% of germinline variants were filtered out for each vcf file
```

(2)	For all the strong tumour neoantigens identified, what are the affected genes? What are the neoantigen peptides? What are the rank scores in tumour and normal, respective, for these strong tumour neoantigens? Create a table for all the strong tumour neoantigens, with all the essential statistics, e.g., including rank scores, and with one column indicating the tumour sample. (7 marks)
```{r}

```

(3)	Are there any shared tumour neoantigens amongst the four tumour samples? What are the potential implications if neoantigen-based therapies are designed? (4 marks)
```{r}

```



# LAQ 4:
MClust tumour clonality anlaysis (Total 17 marks).

In the data folder under the “Assessments” tab, you will find “LU-4.snv.txt” that contains all somatic mutations identified for the LU-4 tumour sample, and “mutations_SP4_LN1_LN2_together.txt” that contains all mutations identified and shared betweel SP4 lymph node 1 (LN1) and lymph node 2 (LN2) tumour samples. This question is to investigate the clonal evolution patterns for a single sample “LU-4”, and for comparing two sites, LN1 vs. LN2, of the SP4 patient, using the mClust analysis.

For the single sample analysis,

(1)	Load the “LU-4.snv.txt” file, and run the mClust single sample clustering analysis using variant allele frequency (VAF) of all mutations identified. First, generate a histogram of all VAFs, and briefly comment on the VAF distribution (2 marks).


```{r}
## Install mclust R package.
if (!requireNamespace("mclust", quietly = TRUE)) {
  install.packages("mclust", repos = "https://cloud.r-project.org")
}

```

```{r}
## Read in snv tab deliminated VCF file “LU-4.snv.txt" provided in the coursework data folder.
## In these files, you can see the read counts for alternative and reference alleles are provided. VAF would be "alternative allele count / (alternative allele count + reference allele count)".

## Load mclust package
library(mclust)

## Set path
setwd("/Users/naomi/Documents/MSc-Cancer-Genomics-DataScience/Module04-CAN7032-and-CAN7132-Computational-Genomics-Transcriptomics-and-Evolution/notebooks/Coursework/data_required_for_CAN7032:7132/LAQ4_data")

## Read in file
data1 <- read.delim("LU-4.snv.txt", header=T)
dim(data1)
head(data1)

### now we calcuate VAF variant allele frequency, as "alternative allele count / (alternative allele count + reference allele count)"
vaf1 <- data1$t_alt_count / (data1$t_alt_count + data1$t_ref_count)
data1$vaf <- vaf1
```
run mclust analysis, and determine the best likelihood of clusters, and estimate the mean of each cluster
```{r}
hist(data1$vaf) ## generate a histogram of VAFs in the tumour sample
```

EXPLAIN OBSERVATIONS OF VAF DISTRIBUTIONS:

The VAF distribution is bimodal and right-skewed, characterized by two distinct clusters of genomic variants. The highest peak on the y-axis occurs at a low frequency (0.05–0.10); while this bin contains the largest number of unique mutations, they represent subclonal or late-occurring variants present in only a small fraction of cells. In contrast, the broader peak between 0.25 and 0.30 represents the clonal population. Even though there are fewer total mutations in this range, their higher VAF indicates they are the dominant mutations present throughout the majority of the tumor mass. The scarcity of variants approaching a VAF of 1.0 suggests the sample is primarily heterozygous and lacks a dominant homozygous population, while the placement of the clonal peak below 0.5 points toward a mix of tumor and normal cells. However, you would need to run mclust mixture modelling to see the actual number of clusters in the tumour sample. When computing the mixture model with mclust, it was shown to have 4 clusters rather than 2 depicted here in the histogram. 


(2)	Run the density estimation analysis via Gaussian finite mixture modelling, followed by dimension reduction clustering and classification analysis. How many clusters can you identify? What is the size (# of mutations) for each cluster? What is the mean VAF value for each cluster? Can you identify which are major clones (heterozygous or homozygous), which are subclones, and their associated cancer cell fractions (CCF)? You can summarise all these into a table as shown in the lecture/practical slides (3 marks).
```{r}
### Dimension reduction for model-based clustering and classification 
### The best number of density cluster can then be estimated based on the Gaussian finite mixture modeling
mod1 <- densityMclust(data1$vaf)
summary(mod1)
plot(mod1, what = "BIC")
plot(mod1, what = "density", data = data1$vaf)
plot(mod1, what = "diagnostic", type = "cdf")
plot(mod1, what = "diagnostic", type = "qq")

mod1dr <- MclustDR(mod1)
summary(mod1dr)
plot(mod1, what = "density", data = data1$vaf, breaks=50)
plot(mod1dr, what="classification")
plot(mod1dr, what="density")
plot(mod1dr, what="evalues")
```

Further manipulate the data of cluster classification.  
```{r}
 mod1dr

### determine the cluster membership for each SNV
mod1dr$classification
mod1dr$class2mixcomp
summary(mod1dr)
### determine the mean value of each cluster according the multi-mixture module, thus the best VAF estimate of each cluster
mod1dr$mu

data1$classification <- mod1dr$classification
```
How many clusters can you identify? 4
What is the size (# of mutations) for each cluster? cluster 1 - 624 mutations, cluster 2 - 595 mutations, cluster 3 - 2823 mutations, cluster 4 - 165 mutations
What is the mean VAF value for each cluster? cluster 1 [1,1]: 0.05827427, cluster 2 [1,2]: 0.08934949, cluster 3 [1,3]: 0.2421108, cluster 4 [1,4]: 0.4471721
- answered by mixture model mclust 

Can you identify which are major clones (heterozygous or homozygous) - cluster 4 (heterozygous as mean VAF is very close to 0.5).
which are subclones - cluster 1, 2, and 3 (all heterozygous as mean VAF below 0.5)
and their associated cancer cell fractions (CCF)? 
first, calculate tumour purity using clonal population. 

cluster 4 CCF (clonal)
Mean VAF: 0.4472328
tumour purity calcualted with clonal mean VAF: 2 x meanVAF of clonal population (cluster 4): 2 x 0.4472328 = 0.8944656 = 89.4% pure sample
now calculate CCF for cluster 4: CCF = (VAFcluster x 2) / purity = 0.4472328 x 2 / 0.8944656 = 1.0

cluster 3 CCF (subclonal)
mean VAF: 0.2421094
CCF  for cluster 3: (VAFcluster x 2) / purity = (0.2421094 x 2) / 0.8944656 = 0.5413498294 = 0.54

cluster 2 CCF (subclonal)
mean VAF: 0.08933489
CCF for cluster 2: CCF = (VAFcluster x 2) / purity = 0.08933489 X 2 / 0.8944656 = 0.199750309 = 0.20

cluster 1 CCF (subclonal)
mean VAF: 0.05826725
CCF for cluster 1: (VAFcluster x 2) / purity = (0.05826725 x 2) / 0.8944656 = 0.1302839371 = 0.13

```{r}
# Cluster Data
results <- data.frame(
  Cluster = c("Cluster 4", "Cluster 3", "Cluster 2", "Cluster 1"),
  Mutations = c(165, 2823, 595, 624),
  Mean_VAF = c(0.4471721, 0.2421108, 0.08934949, 0.05827427)
)

# Define Clonal Mean and Purity
clonal_mean <- results$Mean_VAF[results$Cluster == "Cluster 4"]
purity <- clonal_mean * 2

# Calculate CCF and determine Status
results$CCF <- round((results$Mean_VAF * 2) / purity, 2)
results$Zygosity <- "Heterozygous"
results$Type <- ifelse(results$CCF == 1, "Major Clone", "Subclone")

# Print Table
print(results)

```


(3)	Finally, please draw a tumour evolutionary tree to illustrate the clustering patterns listed above. Within the tree, specifiy clearly the cluster names/numbers and asscoated CCF values (2 marks). ################### CHECK THE ACTUAL WAY TO DO THIS BY JUN AND CHECK THE BRANCHES ARE CORRECT ###############################
```{r}
# Load the igraph library
if(!require(igraph)) install.packages("igraph")
library(igraph)

# Define the relationship (edges)
# Logic: Cluster 4 is the trunk. 3 grows from 4. 2 and 1 grow from 3.
tree_edges <- c(
  "Cluster 4\n(CCF: 1.00)", "Cluster 3\n(CCF: 0.54)",
  "Cluster 3\n(CCF: 0.54)", "Cluster 2\n(CCF: 0.20)",
  "Cluster 3\n(CCF: 0.54)", "Cluster 1\n(CCF: 0.13)"
)

# Create the graph object
tumor_tree <- graph(edges = tree_edges, directed = TRUE)

# Plot the tree
plot(tumor_tree,
     layout = layout_as_tree(tumor_tree, root = 1),
     vertex.size = 60,
     vertex.color = c("#A020F0", "#009E73", "#D55E00", "#56B4E9"), # Match your density plot colors
     vertex.label.color = "white",
     vertex.label.font = 2,
     vertex.label.cex = 0.8,
     edge.arrow.size = 0.6,
     edge.color = "grey60",
     main = "Tumor Evolutionary Tree",
     sub = "Derived from mixture model VAF clusters")
```


For the paired sample analysis,

(4)	Load the “mutations_SP4_LN1_LN2_together.txt” file, and run the mClust paired sample analysis using VAFs of all shared mutations between SP4 LN1 and LN2 samples. First, generate a scatter plot of all VAFs between LN1 and LN2 samples, and briefly comment on the plot on shared mutations between them (2 marks).
```{r}
setwd("data_required_for_CAN7032:7132/LAQ4_data")
getwd()
list.files()

data2 <- read.table("mutations_SP4_LN1_LN2_together.txt", header=T, sep="\t")
data2_vaf <- data.frame(data2$VAF_LN1, data2$VAF_LN2)

### first lets simply look at how the VAFs from LN1 and VAFs from LN2 are correlated by using scatter plot
plot(data2$VAF_LN1, data2$VAF_LN2)

```
(5)	Run mClust analysis based on Gaussian finite mixture model fitted by EM algorithm, followed by dimension reduction clustering and classification analysis. How many clusters can you identify? What is the size (# of mutations) for each cluster? What is the mean VAF value for each cluster in LN1 and LN2 samples, respectively? Can you identify which are major clones (heterozygous or homozygous), which are subclones, and their associated cancer cell fractions (CCF) in LN1 and LN2 samples, respectively? You can summarise all these into a table as shown in the lecture/practical slides (4 marks).

```{r}

### determine Bayesian Information Criterion (BIC) for the clustering performance
BIC <- mclustBIC(data2_vaf)
plot(BIC)

summary(BIC)

mod2 <- Mclust(data2_vaf, x = BIC)
summary(mod2, parameters = TRUE)

###table(class, mod2$classification)

### Dimension reduction for model-based clustering and classification
mod2dr <- MclustDR(mod2)
summary(mod2dr)

### plot the derived number of clusters as scatter plot with different colours for individual clusters
plot(mod2, what = "classification")


### Further manipulate the data and identify cluster membership for each mutation, and determine the mean VAFs for each cluster between the two correlated samples

### determine the cluster membership for each SNV
mod2dr$classification
mod2dr$class2mixcomp
summary(mod2dr)
### determine the mean value of each cluster according the multi-mixture module, thus the best VAF estimate of each cluster
mod2dr$mu

data2$classification <- mod2dr$classification

```


(6)	Finally, please draw a tumour evolutionary tree to illustrate the clustering patterns listed above. Within the tree, specifiy clearly the cluster names/numbers and asscoated CCF values in LN1 and LN2 samples, respectively when applicable (4 marks).


```{r}
 ######################## WHAST THE HELL DOES THIS MEAN??????????????????????/ SORT THIS ###########################################
```

